# 2012 Olympics Swimming - 100m Butterfly Men Finals prediction

*Author: Matt Malin*

Inspired by [mages' blog](http://lamages.blogspot.co.uk/2012/07/london-olympics-and-prediction-for-100m.html)
with predictions for 100m running times, I've decided to perform some basic 
modelling (loess and linear modelling) on previous Olympic results for the 100m 
Butterfly Men's medal winning results.

## Code setup
```{r setup, tidy = FALSE}
library(XML)
library(ggplot2)

swimming_path <- "http://www.databasesports.com/olympics/sport/sportevent.htm?sp=SWI&enum=200"
swimming_path <- "c:/matt/temp/actuarially/Swimming Olympic Medal Winners.htm"

swimming_data <- readHTMLTable(
  readLines(swimming_path), 
  which = 3, 
  stringsAsFactors = FALSE)

# due to some potential errors in passing header = TRUE:
names(swimming_data) <- swimming_data[1, ]
swimming_data <- swimming_data[-1, ]

swimming_data[["Result"]] <- as.numeric(swimming_data[["Result"]])
swimming_data[["Year"]]   <- as.numeric(swimming_data[["Year"]])
swimming_data             <- na.omit(swimming_data)

loess_prediction <- function(
  medal_type = "GOLD", 
  prediction_year = 2012) 
{
  medal_type <- toupper(medal_type)
 
 swimming_loess <- loess(
    Result ~ Year, 
    subset(swimming_data, Medal == medal_type),
    control = loess.control(surface = "direct"))
  
  swimming_prediction <- predict(
    swimming_loess, 
	data.frame(Year = prediction_year), 
	se = FALSE)

  return(swimming_prediction)
}

log_lm_prediction <- function(
  medal_type = "GOLD", 
  prediction_year = 2012) 
{
  medal_type <- toupper(medal_type)
  swimming_log_lm <- lm(
    log(Result) ~ Year, 
    subset(swimming_data, Medal == medal_type))
  
  swimming_prediction <- exp(predict(
    swimming_log_lm, 
	data.frame(Year = prediction_year), 
	se = FALSE))

  return(swimming_prediction)
}

swimming_data <- rbind(
  data.frame(
    swimming_data[c("Year", "Medal", "Result")], 
	type = "actual"),
  data.frame(
    Year = rep(2012, 3),
	Medal = c("GOLD", "SILVER", "BRONZE"),
	Result = c(
	  loess_prediction("gold"), 
	  loess_prediction("silver"),
	  loess_prediction("bronze")),
	type = rep("loess_prediction", 3)))

medal_colours <- c(
  GOLD   = rgb(201, 137, 16, maxColorValue = 255),
  SILVER = rgb(168, 168, 168, maxColorValue = 255),
  BRONZE = rgb(150, 90, 56, maxColorValue = 255))
		
swimming_plot <- ggplot(
  swimming_data,
  aes(
    x = Year, 
	y = Result, 
	colour = Medal, 
	group = Medal)) + 
  scale_x_continuous(limits = c(1968, 2012)) +
  geom_point() + 
  stat_smooth(
    aes(fill = Medal), 
	alpha = 0.25, 
	data = subset(swimming_data, type = "actual"), 
	fullrange = FALSE, 
	method = loess)
	
swimming_plot <- swimming_plot + 
  scale_fill_manual(values = medal_colours) + 
  scale_colour_manual(values = medal_colours) + theme_bw()
```

## Predictions
I now use the functions loess_prediction and log_lm_prediction to estimate the
times for the medal winning times.

### Loess predictions
The gold prediction for 2012 is `r format(loess_prediction("gold"), digits = 3, nsmall = 1)` seconds, for silver is `r format(loess_prediction("silver"), digits = 3, nsmall = 1)` seconds, and for bronze is `r format(loess_prediction("bronze"), digits = 3, nsmall = 1)` seconds.

### Linear modelling (of log results)
I've shown the code here for the calls to the linear modelling approach:

```{r tidy = FALSE, echo = TRUE}
swimming_log_lm_gold   <- log_lm_prediction("gold")
swimming_log_lm_silver <- log_lm_prediction("silver")
swimming_log_lm_bronze <- log_lm_prediction("bronze")
```

This gives the following times as predictions:

```{r log_lm_predictions}
swimming_log_lm_gold
swimming_log_lm_silver
swimming_log_lm_bronze
```

### Loess prediction plot
The following is a plot of actual and predicted times, along with loess error
setting as defaults from geom_smooth:

```{r plot, echo = FALSE}
swimming_plot
```

## Notes

Note that because of the small difference between the silver and gold medal 
results at the 2008 olympics, the trend of improvement in silver exceeds that
in the gold, so the prediction is that the silver time will be faster than the 
gold!

Also note that this takes into account no information about performance of 
athletes involved or changes in rules, such as being unable to use the swimsuits
that were present in the last Olympics and largely attributed to improving 
performance, purely modelling from a few data points as an interesting exercise!

## Final Summary

To summarise, the final predicted results using these methods are:
```{r final_summary, echo = TRUE, tidy = FALSE, eval = TRUE, results = "asis"}
library(pander)
predictions <- data.frame(
  Medal = c("Gold", "Silver", "Bronze"),
  Loess_prediction = c(
    loess_prediction("gold"),
    loess_prediction("silver"),
    loess_prediction("bronze")),
  Log_Linear_prediction = c(
    log_lm_prediction("gold"),
    log_lm_prediction("silver"),
    log_lm_prediction("bronze")))
pandoc.table(predictions)
```

Now let's see what actually happens, and if it'll be Michael Phelps yet again!